Automated Formal Synthesis of Digital Controllers for State-Space Physical Plants

### 2017-01-21 cav2017/cav2017.tex
Fixing spelling mistakes

\begin{abstract}
This work presents a sound and automated approach to synthesise 
digital feedback control architectures for physical plants represented 
as linear, time invariant models. Models are encompassed by dynamical 
equations with inputs, evolving over a continuous state space describing 
the evolution of the physical variables of the system.  The approach is divided 
in two main parts, brought together by a counter-example guided inductive 
synthesis (CEGIS) loop, as follows. First, we devise a static feedback controller 
that stabilises the system; then, we verify its potential safety by means of either 
of the following approaches: forward reachability analysis via unfolding of the dynamics, 
or invariant set generation via abstract acceleration;  
further, if the above step fails, we loop back and invoke a CEGIS call 
to seek alternative stabilising feedbacks, benefitting from the obtained counter-examples 
to safety. 
\textcolor{red}{[Mention GA \todo{GA? Genetic Algorithms? We don't use GA?} and Abstract Acceleration here?]}

@@

Linear Time Invariant (LTI) control models represent a broad class of dynamical systems
%modern embedded software, performing highly non-trivial control tasks,
with significant impact in numerous application areas such as
the life sciences, robotics, and engineering~\cite{astrom1997computer,Franklin15}.
Control synthesis for LTI is a broadly investigated, 
however the presence of digital control architectures is exacerbated by specific artefacts,  
%specific to digital control, 
such as the effects of finite-precision arithmetic, time discretisation, and quantisation noise, which are typically introduced 
by Analog-to-Digital (A/D) and Digital-to-Analog (D/A) conversion.  
Whilst the research on digital control is well developed [Astrom book on digital control], 
automated and sound control synthesis is a challenging task. 
%Thus, the programming is a key barrier to broad adoption of digital control, 
%and requires considerable expertise.

@@

There are two different problems that arise from FWL semantics.  The first
is the error in the dynamics caused by the inability to represent the exact
state of the physical system while the second relates to rounding errors
during computation.  In~\cite{fialho1994stability}, a stability measure
based on the error of the digital dynamics ensures that the deviation
introduced by FWL does not make the digital system unstable.  A more recent
approach~\cite{DBLP:journals/automatica/WuLCC09} uses $\mu$ calculus to
directly model the digital controller so that the selected parameters are
stable by design.  The analyses
in~\cite{DBLP:conf/hybrid/RouxJG15,DBLP:conf/hybrid/WangGRJF16} rely on an
invariant computation on the discrete system dynamics using Semi-Definite
Programming (SDP).  While the former uses BIBO properties to determine
stability, the latter uses Lyapunov-based quadratic invariants.  In both
cases, the SDP solver uses floating-point arithmetic and soundness is
checked by bounding the error.  An alternative approach is taken
by~\cite{park2016scalable}, where the verification of existing code is
performed against a known model by extracting an LTI model of the code
through symbolic execution.  In order to account for rounding errors, an
upper bound is introduced in the verification phase.
The work in \cite{picasso2003stabilization,picasso2002construction} introduces
attractive invariant sets as a mechanism to bound the quantization error effect
on stabilization as an invariant set that always converges toward the cotrollable
set. In a similar fashion, \cite{liberzon2003hybrid} evaluates the dynamics of the
quantization error and binds their trajectory to a known region over a finite
period of time. This technique works for both linear and non-linear systems.
This paper uses an approach that projects the maximum effect of the noise over
an unbounded time and uses it as a `buffer' region for the safety specification.

@@

We consider physical models expressed as ordinary differential equations (ODEs) with full state information:  
%
\begin{align}
\label{eq:ode}
\dot{x}(t) = Ax(t)+ B u(t)\ \ :\ \ x \in \mathbb{R}^{n}, u \in \mathbb{R}^m, A \in \mathbb{R}^{n \times n}, B \in \mathbb{R}^{n \times m}
\end{align}
where $A$ and $B$ are matrices that fully specify the continuous plant with the initial states $x(0)$. 
Eq.~\eqref{eq:ode} is soundly discretised (as shown in the Appendix) into
%
\begin{align}
\label{eq:plant}
x_{k+1} = A_d x_k+ B_d u_k. 
\end{align}
with initial state $x_{0}=x(0)$.

@@

%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Solution generalisation} 
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
Given the large soloution space to be searched during the inductive
synthesis phase, we parametrise the target language, which induces a
lattice of progressively more expressive languages.  This enables us
to start by attempting to synthesise a program at the lowest point on
this lattice and increase the parameters of the language until we
reach a point at which the synthesis succeeds. Whenever we manage to
synthesise a candidate solution at some lower point on the lattice, we
must apply some form of solution generalisation to obtain a full
solution (i.e. a solution in the original search space).

@@

We investigate two different approaches of integrating solution generalisation 
with CEGIS:
\begin{itemize}
\item[1.] The first approach uses a multi-staged validation phase
  and performes solution generalisation based on
the plant's precision, the number of loop unrollings
and the sampling rate (see Sec.~\ref{sec:CEGIS-precision-incrementation}).
\item[2.] The second approach uses abstract acceleration to compute the set of reachable states,
  which enables it to perform abstraction refinement inside the CEGIS loop (see Sec~\ref{sec:CEGIS-abstraction-refinement}).

@@

\addtodo{Revise: }
The first aproach finds a controller for an individual initial state and input with a bounded time horizon and generalizes to all states, inputs and time horizons (with a final verification loop that verifies the continuous system behaviour against the previously verified discrete system bisimulation).
The second approach finds a controller for a continuous initial state-space and input sets over an abstraction of the continuous dynamics which conforms to witness proofs at specific times k, and generalizes the solution for all times by refining the number of witnesses.

@@

The verification of a hybrid continuous/discrete time/space system over an unbounded time, requires the
non-deterministic simulation of an infinite set of states, which is undecidable. Our initial approach creates an
abstraction of this system by establishing bounds on the error between such a system and a discrete-time/space model.
This model is still however, quite complex, hard to verify, and many of the states involved are inconsequential to
the safety specification.
Our second approach, rather than look at a direct bisimulation in discrete time, explores a much higher abstraction,
evaluating the unbounded time-continuous space reach tube as a set of inequalities that represent only the safety specifcation.
We begin with a minimum set of linear inequalities describing the constraints caused by the input and the initial state during a
single iteration, in addition to a stability specification defined by Jury's criteria, and try to synthesise a controller that meets these specs.
A second abstraction that encompasses the accelerated continuous reach-tube over an unbounded time is used to verify that the controller meets the full safety specification. If this step fails, we find a counterexmple iteration that causes the selected system to fail and create a new constraint to refine our abstraction.
The advantage of this model is that it starts with a very simple description regardless of the complexity of the overall dynamics, and only expands to more complex models when a solution cannot be found.

@@

The characteristic polynomial of the above matix is:
%
\begin{align*}
z^n+\sum_{i=1}^p{(a_i-k_i)z^{n-i}}
\end{align*}

@@

Discretization of an continuous dynamical system turns the ODE into a difference equation, assuming zero-order hold
for the input $\vec{u}$ and continuous integration for the noise $\vec{w}$, to
\begin{align}
\label{eq:discretization}
\vec{x}_{k+1} &= \vec{A}_d\vec{x}_k+\vec{B}_d\vec{u}_k + \vec{w}_k,\\
y_k &= \vec{C}_d \vec{x}_ k + \vec{D}_d \vec{u}_ k. 
\end{align}
with covariance for $\vec{w}_k \sim \mathcal{N}(0,Q_d)$,
where
\begin{align}
\label{eq:discretize}
\vec{A}_d &= e^{\vec{A} T_s} = \mathcal{L}^{-1} { ( s \vec{I} - \vec{A} )^{-1} }_{t = T_s},\\
\vec{B}_d &= \int_{0}^{T_s} e^{\vec{A} t} dt\ \vec{B} = \vec{A}^{-1} ( \vec{A}_d - \vec{I} ) \vec{B},\\
\vec{C}_d &= \vec{C},\\
\vec{D}_d &= \vec{D},\\
Q_d &= \int_{0}^{T_s} e^{\vec{A} t} Q e^{\vec{A}^T t} dt.
\end{align}
and $T_s$ is the sample time. Then
\begin{align*}
x(kT)=x_k \wedge y(kT) = y_k, \forall k.
\end{align*}
Let $g_d(k)=\mathcal{G}(t,g(t),T)$ be a function that performs the discretization described above, where $g(t)$
represents the continuous dynamics and $g_d(k)$ the corresponding discrete dynamics. 
Let $G(s)=\mathcal{L}(g(t))$ and $G_d(z)=\mathcal{Z}(g_d(k))$ be the corresponding Laplace and Z-transforms
of $g(t)$ and $g_d(k)$. Given this relation, we have 
$$G_d(z)=G(z)|_{z=e^{sT}} : g_d(k)=\mathcal{G}(t,g(t),T) \wedge T < \frac{1}{.5f_s},$$
where $.5f_s$ is the Nyquist frequency of $g(t)$. This last restriction is introduced to avoid the effects of aliasing
which could cause ``phantom poles'' to appear otherwise. 
The eigenvalues of $\vec{A}_d$ corresponding to the poles of $G_d(z)$, and those of $\vec{A}$ corresponding to the poles of $G(s)$ are similarly related.
$\hat{\lambda}_i=e^{-\lambda_iT} : \hat{\lambda}_i \in \sigma(\vec{A}_d) \wedge \lambda_i \in \sigma(\vec{A})$
where $\sigma(\cdot)$ is the spectrum of a matrix.
The following remark is worth mentioning regarding the above.
\begin{remark}
The witnessed maximum amplitude of the discrete signals $x_k,y_k$ may be smaller to that of $x(t),y(t)$ due to synchronism at fraction-frequency sampling.
This means that reasoning about the state-space must consider the effects of this possibility as well of the case of maximal inputs from the continuous specification.
\end{remark}
 Since \eqref{eq:discretization} is a bisimulation of \eqref{eq:ode}, we may use the semantics in section \ref{sec:model_semantics} to model continuous dynamical systems.

@@

During any given ADC conversion, the continuous signal will be sampled in the real domain and transformed into a $\mathbb{R}\langle I_{adc},F_{adc} \rangle$ value. this sampling uses a threshold which is defined by the less significant bit ($q_{adc}=2^{-F_{adc}}$) of the adc and some nonlinearities of the circuitry. The overall conversion is
$$\mathcal{F}\langle I_{adc},F_{adc} \rangle(y(t)) = y_k : y_k \in \left[y(t)-\frac{q_{adc}}{2}\ \ \ \ y(t)+\frac{q_{adc}}{2}\right]$$.
If we denote the error in the conversion as $\nu_k=y_k-y(t) : t=nk$ then we may define some bounds for it $\nu_k \in [-\frac{q_{adc}}{2}\ \ \frac{q_{adc}}{2}]$.

We will assume, for the purposes of this analysis, that the domain of the ADC is that of the digital controller (\emph{ie} the quantizer includes any digital gain added in the code).
The process of quantization in the DAC is similar except that it is going from $\mathbb{R}\langle I_{adc},F_{adc} \rangle \rightarrow \mathbb{R}\langle I_{dac},F_{dac} \rangle$. If these domains are the same ($I_{adc}=I_{dac},F_{adc}=F_{dac}$), or if the DAC resolution in higher than the ADCs, then the DAC quantization error is equal to zero.
From the above equations we can now infer that $\nu_1 \in [-\frac{q_1}{2}\ \ \frac{q_1}{2}], \nu_2 \in [-\frac{q_2}{2}\ \ \frac{q_2}{2}] : q_1=q_{adc} \wedge q_2=q_{dac}$.
Note that these bounds hold irrespective of whether the noise is correlated, hence we may use them to overapproximate the effect of the noise on the state space progression over time.
The resulting dynamics are:
\begin{align}
\label{eq:pre_quantization}
\vec{x}_{k+1} &= \vec{A}_d\vec{x}_k+\vec{B}_d(\vec{u}_k+{\vec{\nu}_2}_k) + \vec{w}_k,\\
y_k &= \vec{C}_d \vec{x}_ k + \vec{D}_d \vec{u}_ k+{\vec{\nu}_1}_k. 
\end{align}

>>>>

\begin{abstract}
This work presents a sound and automated approach to synthesise 
digital feedback control architectures for physical plants represented 
as linear, time invariant models. Models are encompassed by dynamical 
equations with inputs, evolving over a continuous state space describing 
the evolution of the physical variables of the system.  The approach is divided 
in two main parts, brought together by a counter-example guided inductive 
synthesis (CEGIS) loop, as follows. First, we devise a static feedback controller 
that stabilises the system; then, we verify its potential safety by means of either 
of the following approaches: forward reachability analysis via unfolding of the dynamics, 
or invariant set generation via abstract acceleration;  
further, if the above step fails, we loop back and invoke a CEGIS call 
to seek alternative stabilising feedbacks, benefiting from the obtained counter-examples 
to safety. 
\textcolor{red}{[Mention GA \todo{GA? Genetic Algorithms? We don't use GA?} and Abstract Acceleration here?]}

@@

Linear Time Invariant (LTI) control models represent a broad class of dynamical systems
%modern embedded software, performing highly non-trivial control tasks,
with significant impact in numerous application areas such as
the life sciences, robotics, and engineering~\cite{astrom1997computer,Franklin15}.
Control synthesis for LTI is a broadly investigated, 
however the presence of digital control architectures is exacerbated by specific artefacts,  
%specific to digital control, 
such as the effects of finite-precision arithmetic, time discretization, and quantisation noise, which are typically introduced 
by Analogue-to-Digital (A/D) and Digital-to-Analogue (D/A) conversion.  
Whilst the research on digital control is well developed [Astrom book on digital control], 
automated and sound control synthesis is a challenging task. 
%Thus, the programming is a key barrier to broad adoption of digital control, 
%and requires considerable expertise.

@@

There are two different problems that arise from FWL semantics.  The first
is the error in the dynamics caused by the inability to represent the exact
state of the physical system while the second relates to rounding errors
during computation.  In~\cite{fialho1994stability}, a stability measure
based on the error of the digital dynamics ensures that the deviation
introduced by FWL does not make the digital system unstable.  A more recent
approach~\cite{DBLP:journals/automatica/WuLCC09} uses $\mu$ calculus to
directly model the digital controller so that the selected parameters are
stable by design.  The analyses
in~\cite{DBLP:conf/hybrid/RouxJG15,DBLP:conf/hybrid/WangGRJF16} rely on an
invariant computation on the discrete system dynamics using Semi-Definite
Programming (SDP).  While the former uses BIBO properties to determine
stability, the latter uses Lyapunov-based quadratic invariants.  In both
cases, the SDP solver uses floating-point arithmetic and soundness is
checked by bounding the error.  An alternative approach is taken
by~\cite{park2016scalable}, where the verification of existing code is
performed against a known model by extracting an LTI model of the code
through symbolic execution.  In order to account for rounding errors, an
upper bound is introduced in the verification phase.
The work in \cite{picasso2003stabilization,picasso2002construction} introduces
attractive invariant sets as a mechanism to bound the quantization error effect
on stabilization as an invariant set that always converges toward the controllable
set. In a similar fashion, \cite{liberzon2003hybrid} evaluates the dynamics of the
quantization error and binds their trajectory to a known region over a finite
period of time. This technique works for both linear and non-linear systems.
This paper uses an approach that projects the maximum effect of the noise over
an unbounded time and uses it as a `buffer' region for the safety specification.

@@

We consider physical models expressed as ordinary differential equations (ODEs) with full state information:  
%
\begin{align}
\label{eq:ode}
\dot{x}(t) = Ax(t)+ B u(t)\ \ :\ \ x \in \mathbb{R}^{n}, u \in \mathbb{R}^m, A \in \mathbb{R}^{n \times n}, B \in \mathbb{R}^{n \times m}
\end{align}
where $A$ and $B$ are matrices that fully specify the continuous plant with the initial states $x(0)$. 
Eq.~\eqref{eq:ode} is soundly discretized (as shown in the Appendix) into
%
\begin{align}
\label{eq:plant}
x_{k+1} = A_d x_k+ B_d u_k. 
\end{align}
with initial state $x_{0}=x(0)$.

@@

%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Solution generalisation} 
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
Given the large solution space to be searched during the inductive
synthesis phase, we parametrise the target language, which induces a
lattice of progressively more expressive languages.  This enables us
to start by attempting to synthesise a program at the lowest point on
this lattice and increase the parameters of the language until we
reach a point at which the synthesis succeeds. Whenever we manage to
synthesise a candidate solution at some lower point on the lattice, we
must apply some form of solution generalisation to obtain a full
solution (i.e. a solution in the original search space).

@@

We investigate two different approaches of integrating solution generalisation 
with CEGIS:
\begin{itemize}
\item[1.] The first approach uses a multi-staged validation phase
  and performs solution generalisation based on
the plant's precision, the number of loop unrollings
and the sampling rate (see Sec.~\ref{sec:CEGIS-precision-incrementation}).
\item[2.] The second approach uses abstract acceleration to compute the set of reachable states,
  which enables it to perform abstraction refinement inside the CEGIS loop (see Sec~\ref{sec:CEGIS-abstraction-refinement}).

@@

\addtodo{Revise: }
The first approach finds a controller for an individual initial state and input with a bounded time horizon and generalizes to all states, inputs and time horizons (with a final verification loop that verifies the continuous system behaviour against the previously verified discrete system bi-simulation).
The second approach finds a controller for a continuous initial state-space and input sets over an abstraction of the continuous dynamics which conforms to witness proofs at specific times k, and generalizes the solution for all times by refining the number of witnesses.

@@

The verification of a hybrid continuous/discrete time/space system over an unbounded time, requires the
non-deterministic simulation of an infinite set of states, which is undecidable. Our initial approach creates an
abstraction of this system by establishing bounds on the error between such a system and a discrete-time/space model.
This model is still however, quite complex, hard to verify, and many of the states involved are inconsequential to
the safety specification.
Our second approach, rather than look at a direct bi-simulation in discrete time, explores a much higher abstraction,
evaluating the unbounded time-continuous space reach tube as a set of inequalities that represent only the safety specification.
We begin with a minimum set of linear inequalities describing the constraints caused by the input and the initial state during a
single iteration, in addition to a stability specification defined by Jury's criteria, and try to synthesise a controller that meets these specifications.
A second abstraction that encompasses the accelerated continuous reach-tube over an unbounded time is used to verify that the controller meets the full safety specification. If this step fails, we find a counterexample iteration that causes the selected system to fail and create a new constraint to refine our abstraction.
The advantage of this model is that it starts with a very simple description regardless of the complexity of the overall dynamics, and only expands to more complex models when a solution cannot be found.

@@

The characteristic polynomial of the above matrix is:
%
\begin{align*}
z^n+\sum_{i=1}^p{(a_i-k_i)z^{n-i}}
\end{align*}

@@

Discretization of an continuous dynamical system turns the ODE into a difference equation, assuming zero-order hold
for the input $\vec{u}$ and continuous integration for the noise $\vec{w}$, to
\begin{align}
\label{eq:discretization}
\vec{x}_{k+1} &= \vec{A}_d\vec{x}_k+\vec{B}_d\vec{u}_k + \vec{w}_k,\\
y_k &= \vec{C}_d \vec{x}_ k + \vec{D}_d \vec{u}_ k. 
\end{align}
with covariance for $\vec{w}_k \sim \mathcal{N}(0,Q_d)$,
where
\begin{align}
\label{eq:discretize}
\vec{A}_d &= e^{\vec{A} T_s} = \mathcal{L}^{-1} { ( s \vec{I} - \vec{A} )^{-1} }_{t = T_s},\\
\vec{B}_d &= \int_{0}^{T_s} e^{\vec{A} t} dt\ \vec{B} = \vec{A}^{-1} ( \vec{A}_d - \vec{I} ) \vec{B},\\
\vec{C}_d &= \vec{C},\\
\vec{D}_d &= \vec{D},\\
Q_d &= \int_{0}^{T_s} e^{\vec{A} t} Q e^{\vec{A}^T t} dt.
\end{align}
and $T_s$ is the sample time. Then
\begin{align*}
x(kT)=x_k \wedge y(kT) = y_k, \forall k.
\end{align*}
Let $g_d(k)=\mathcal{G}(t,g(t),T)$ be a function that performs the discretization described above, where $g(t)$
represents the continuous dynamics and $g_d(k)$ the corresponding discrete dynamics. 
Let $G(s)=\mathcal{L}(g(t))$ and $G_d(z)=\mathcal{Z}(g_d(k))$ be the corresponding Laplace and Z-transforms
of $g(t)$ and $g_d(k)$. Given this relation, we have 
$$G_d(z)=G(z)|_{z=e^{sT}} : g_d(k)=\mathcal{G}(t,g(t),T) \wedge T < \frac{1}{.5f_s},$$
where $.5f_s$ is the Nyquist frequency of $g(t)$. This last restriction is introduced to avoid the effects of aliasing
which could cause ``phantom poles'' to appear otherwise. 
The eigenvalues of $\vec{A}_d$ corresponding to the poles of $G_d(z)$, and those of $\vec{A}$ corresponding to the poles of $G(s)$ are similarly related.
$\hat{\lambda}_i=e^{-\lambda_iT} : \hat{\lambda}_i \in \sigma(\vec{A}_d) \wedge \lambda_i \in \sigma(\vec{A})$
where $\sigma(\cdot)$ is the spectrum of a matrix.
The following remark is worth mentioning regarding the above.
\begin{remark}
The witnessed maximum amplitude of the discrete signals $x_k,y_k$ may be smaller to that of $x(t),y(t)$ due to synchronism at fraction-frequency sampling.
This means that reasoning about the state-space must consider the effects of this possibility as well of the case of maximal inputs from the continuous specification.
\end{remark}
 Since \eqref{eq:discretization} is a bi-simulation of \eqref{eq:ode}, we may use the semantics in section \ref{sec:model_semantics} to model continuous dynamical systems.

@@

During any given ADC conversion, the continuous signal will be sampled in the real domain and transformed into a $\mathbb{R}\langle I_{adc},F_{adc} \rangle$ value. this sampling uses a threshold which is defined by the less significant bit ($q_{adc}=2^{-F_{adc}}$) of the ADC and some non-linearities of the circuitry. The overall conversion is
$$\mathcal{F}\langle I_{adc},F_{adc} \rangle(y(t)) = y_k : y_k \in \left[y(t)-\frac{q_{adc}}{2}\ \ \ \ y(t)+\frac{q_{adc}}{2}\right]$$.
If we denote the error in the conversion as $\nu_k=y_k-y(t) : t=nk$ then we may define some bounds for it $\nu_k \in [-\frac{q_{adc}}{2}\ \ \frac{q_{adc}}{2}]$.

We will assume, for the purposes of this analysis, that the domain of the ADC is that of the digital controller (\emph{ie} the quantizer includes any digital gain added in the code).
The process of quantization in the DAC is similar except that it is going from $\mathbb{R}\langle I_{adc},F_{adc} \rangle \rightarrow \mathbb{R}\langle I_{dac},F_{dac} \rangle$. If these domains are the same ($I_{adc}=I_{dac},F_{adc}=F_{dac}$), or if the DAC resolution in higher than the ADCs, then the DAC quantization error is equal to zero.
From the above equations we can now infer that $\nu_1 \in [-\frac{q_1}{2}\ \ \frac{q_1}{2}], \nu_2 \in [-\frac{q_2}{2}\ \ \frac{q_2}{2}] : q_1=q_{adc} \wedge q_2=q_{dac}$.
Note that these bounds hold irrespective of whether the noise is correlated, hence we may use them to over-approximate the effect of the noise on the state space progression over time.
The resulting dynamics are:
\begin{align}
\label{eq:pre_quantization}
\vec{x}_{k+1} &= \vec{A}_d\vec{x}_k+\vec{B}_d(\vec{u}_k+{\vec{\nu}_2}_k) + \vec{w}_k,\\
y_k &= \vec{C}_d \vec{x}_ k + \vec{D}_d \vec{u}_ k+{\vec{\nu}_1}_k. 
\end{align}
